\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2015-04-24 doc/USER_MANUAL/12_running_scheduler.tex
updates user manual to use facilitate the generation of a markdown version to the github wiki

% !TEX encoding = IsoLatin
%% LyX 2.0.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[oneside,english]{book}
\usepackage[T1]{fontenc}
%% DK DK we do not accept accented letters, use the old LaTeX syntax
%% DK DK for them instead: \'e \`e  \"e \^e and so on.
%% DK DK \usepackage[latin1]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{babel}
\usepackage{float}
\usepackage{textcomp}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=false,
 breaklinks=false,pdfborder={0 0 0},backref=section,colorlinks=false,pdfpagemode=None]
 {hyperref}
\hypersetup{
 pdftex,pdfstartview=FitH,pdfpagelayout=SinglePage}
\usepackage{breakurl}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\newcommand{\noun}[1]{\textsc{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newenvironment{lyxcode}
{\par\begin{list}{}{
\setlength{\rightmargin}{\leftmargin}
\setlength{\listparindent}{0pt}% needed for AMS classes
\raggedright
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}
\normalfont\ttfamily}%
 \item[]}
{\end{list}}
\newenvironment{lyxlist}[1]
{\begin{list}{}
{\settowidth{\labelwidth}{#1}
 \setlength{\leftmargin}{\labelwidth}
 \addtolength{\leftmargin}{\labelsep}
 \renewcommand{\makelabel}[1]{##1\hfil}}}
{\end{list}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
%\renewcommand{\baselinestretch}{1.5}

% figures
\usepackage[dvips]{epsfig}

% fonts
\usepackage{times}

% hyperlinks to sections and references

\newcommand{\urlwithparentheses}[1]{(\url{#1})}

% biblio GJI
\bibliographystyle{abbrvnat}

\newcommand{\nexxi}{\mbox{\texttt{NEX\_XI\/}}}
\newcommand{\nexeta}{\mbox{\texttt{NEX\_ETA\/}}}
\newcommand{\nprocxi}{\mbox{\texttt{NPROC\_XI\/}}}
\newcommand{\nproceta}{\mbox{\texttt{NPROC\_ETA\/}}}
\newcommand{\nchunks}{\mbox{\texttt{NCHUNKS\/}}}
\newcommand{\eg}{e.g.,}

% colors for the text
\@ifundefined{definecolor}
 {\usepackage{color}}{}
\newcommand{\red}[1]{\textcolor{Red}{#1}}

% some of Dimitri's aliases for math symbols
\newcommand{\Grad}{\mbox{\boldmath $\bf \nabla$}}
\newcommand{\Div}{\Grad \cdot }
\newcommand{\Laplacian}{\nabla^2}
\newcommand{\pressure}{p}
\newcommand{\Int}{\displaystyle\int}
\newcommand{\Sum}{\displaystyle\sum}
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\myiomega}{\mathbf{i} \, \omega}
\newcommand{\p}{\partial}
\newcommand{\boldu}{\mathbf{u}}
\newcommand{\boldv}{\mathbf{v}}
\newcommand{\boldw}{\mathbf{w}}
\newcommand{\boldx}{\mathbf{x}}
\newcommand{\boldy}{\mathbf{y}}
\newcommand{\boldm}{\mathbf{m}}
\newcommand{\boldp}{\mathbf{p}}
\newcommand{\boldg}{\mathbf{g}}
\newcommand{\bolds}{\mathbf{s}}
\newcommand{\boldn}{\mathbf{\hat{n}}}
\newcommand{\boldd}{\mathbf{d}}
\newcommand{\boldf}{\mathbf{f}}
\newcommand{\boldB}{\mathbf{B}}
\newcommand{\boldc}{\mathbf{c}}
\newcommand{\boldC}{\mathbf{C}}
\newcommand{\boldE}{\mathbf{E}}
\newcommand{\boldH}{\mathbf{H}}
\newcommand{\boldI}{\mathbf{I}}
\newcommand{\boldsigma}{\mbox{\boldmath $\bf \sigma$}}
\newcommand{\boldxi}{\mbox{\boldmath $\bf \xi$}}
\newcommand{\boldvarepsilon}{\mbox{\boldmath $\bf \varepsilon$}}
\newcommand{\bolddelta}{\mbox{\boldmath $\bf \delta$}}
\newcommand{\boldtau}{\mbox{\boldmath $\bf \tau$}}
\newcommand{\Boxmath}{\Lambda}
\newcommand{\rmd}{\, \mathrm{d}}
\newcommand{\pv}{\ ,}
\newcommand{\pp}{\ .}

\usepackage{babel}

\makeatother

\begin{document}
\begin{center}
\thispagestyle{empty}\vspace*{-1.8truecm} %
\makebox[1\textwidth]{%
\includegraphics[width=0.83\paperwidth]{figures/specfem_3d_Cartesian-cover} %
}
\par\end{center}

\title{\textbf{SPECFEM3D Cartesian}\\
 \textbf{User Manual}}

\author{$\copyright$ Princeton University (USA), CNRS and University of Marseille (France),\\
and ETH Z\"urich (Switzerland)\\
Version 3.0}

\date{\today}

\maketitle

\section*{Authors}

\noindent The SPECFEM3D Cartesian package was first developed by Dimitri
Komatitsch and Jean-Pierre Vilotte at Institut de Physique du Globe
(IPGP) in Paris, France from 1995 to 1997 and then by Dimitri Komatitsch
and Jeroen Tromp at Harvard University and Caltech, USA, starting in 1998.
The story started on April 4, 1995, when Prof. Yvon Maday from CNRS and University of Paris, France, gave a lecture to
Dimitri Komatitsch and Jean-Pierre Vilotte at IPG about the nice properties of the Legendre spectral-element method with diagonal mass matrix that he had used for
other equations. We are deeply indebted and thankful to him for that.
That followed a visit by Dimitri Komatitsch to OGS (Istituto Nazionale di Oceanografia e di Geofisica Sperimentale) in Trieste, Italy, in February 1995
to meet with G\'eza Seriani and Enrico Priolo, who introduced him to their 2D Chebyshev version of the spectral-element method with a non-diagonal mass matrix.
We are deeply indebted and thankful to them for that.

%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
%%%%%%%% please do NOT merge the lines of the paragraph below, it makes it difficult to insert new names in alphabetical order; leave a single name per line
Since then it has been developed and maintained by a development team:
in alphabetical order,
Jean-Paul (Pablo) Ampuero,
Kangchen Bai,
Piero Basini,
C\'eline Blitz,
Alexis Bottero,
Ebru Bozda\u{g},
Emanuele Casarotti,
Joseph Charles,
Min Chen,
Paul Cristini,
Percy Galvez,
Dominik G\"oddeke,
Vala Hj\"orleifsd\'ottir,
Sue Kientz,
Dimitri Komatitsch,
Jes\'us Labarta,
Nicolas Le Goff,
Pieyre Le Loher,
Matthieu Lefebvre,
Qinya Liu,
Yang Luo,
Alessia Maggi,
Federica Magnoni,
Roland Martin,
Ren\'e Matzen,
Dennis McRitchie,
Matthias Meschede,
Peter Messmer,
David Mich\'ea,
Surendra Nadh Somala,
Tarje Nissen-Meyer,
Daniel Peter,
Max Rietmann,
Elliott Sales de Andrade,
Brian Savage,
Bernhard Schuberth,
Anne Sieminski,
Leif Strand,
Carl Tape,
Jeroen Tromp,
Jean-Pierre Vilotte,
Zhinan Xie,
Chang-Hua Zhang,
Hejun Zhu.\\

The cover graphic of the manual was created by Santiago Lombeyda from
Caltech's Center for Advanced Computing Research (CACR) \urlwithparentheses{http://www.cacr.caltech.edu}.\\

The code is released open-source under the GNU version 2+ license, see the license at the end of this manual.\\

\newpage{}
\section*{Current and past main participants or main sponsors of the SPECFEM project (in no particular order)}
%
\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[width=0.125\textwidth]{figures/logo_cnrs}\vspace*{2truemm}
\includegraphics[width=0.125\textwidth]{figures/logo_princeton}\vspace*{2truemm}
\includegraphics[width=0.15\textwidth]{figures/logo_aix_marseille_universite}\vspace*{0.02truemm}
\includegraphics[width=0.15\textwidth]{figures/logo_ETH}\vspace*{2truemm}
\includegraphics[width=0.12\textwidth]{figures/logo_CSC_China}\vspace*{0.02truemm}
\includegraphics[width=0.15\textwidth]{figures/logo_inria}\vspace*{2truemm}
\includegraphics[width=0.14\textwidth]{figures/logo_UPPA}
\par\end{centering}
%
\vspace*{-2truemm}
%
\noindent \begin{centering}
\includegraphics[width=0.15\textwidth]{figures/logo_NVIDIA}\vspace*{2truemm}
\includegraphics[width=0.125\textwidth]{figures/logo_IUF}\vspace*{2truemm}
\includegraphics[width=0.135\textwidth]{figures/logo_Caltech}\vspace*{2truemm}
\includegraphics[width=0.135\textwidth]{figures/logo_Harvard}\vspace*{2truemm}
\includegraphics[width=0.135\textwidth]{figures/logo_IPGP}\vspace*{2truemm}
\includegraphics[width=0.112\textwidth]{figures/logo_ANR}\vspace*{2truemm}
\includegraphics[width=0.112\textwidth]{figures/logo_NSF}\vspace*{2truemm}
\par\end{centering}
%
\vspace*{-2truemm}
%
\noindent \begin{centering}
\includegraphics[width=0.190\textwidth]{figures/logo_European_Union}\vspace*{2truemm}
\includegraphics[width=0.112\textwidth]{figures/logo_GENCI}\vspace*{2truemm}
\includegraphics[width=0.112\textwidth]{figures/logo_PRACE}\vspace*{2truemm}
\includegraphics[width=0.112\textwidth]{figures/logo_CINES}\vspace*{2truemm}
\includegraphics[width=0.130\textwidth]{figures/logo_Oak_Ridge}\vspace*{2truemm}
\hspace*{3mm}\includegraphics[width=0.130\textwidth]{figures/logo_fondation_Del_Duca}
\includegraphics[width=0.130\textwidth]{figures/logo_CIG}\vspace*{2truemm}
\par\end{centering}
%
\vspace*{-2truemm}
%
\noindent \begin{centering}
\includegraphics[width=0.112\textwidth]{figures/logo_University_of_Toronto}\vspace*{2truemm}
\includegraphics[width=0.109\textwidth]{figures/logo_INGV}\vspace*{2truemm}
\includegraphics[width=0.102\textwidth]{figures/logo_Univ_Toulouse}\vspace*{2truemm}
\includegraphics[width=0.112\textwidth]{figures/logo_TOTAL}\vspace*{2truemm}
\includegraphics[width=0.130\textwidth]{figures/logo_Fairbanks}\vspace*{2truemm}
\includegraphics[width=0.120\textwidth]{figures/logo_CINECA}\vspace*{2truemm}
\includegraphics[width=0.140\textwidth]{figures/logo_Intel_Exascale_Labs}\vspace*{2truemm}
\includegraphics[width=0.130\textwidth]{figures/logo_Maison_Simulation}
\par\end{centering}
\end{figure}

\newpage{}

\tableofcontents{}

\chapter{Introduction}

The software package SPECFEM3D Cartesian simulates seismic wave propagation
at the local or regional scale and performs full waveform imaging (FWI) or adjoint tomography based upon the spectral-element method
(SEM). The SEM is a continuous Galerkin technique \citep{TrKoLi08,PeKoLuMaLeCaLeMaLiBlNiBaTr11},
which can easily be made discontinuous \citep{BeMaPa94,Ch00,KoWoHu02,ChCaVi03,LaWaBe05,Kop06,WiStBuGh10,AcKo11};
it is then close to a particular case of the discontinuous Galerkin
technique \citep{ReHi73,LeRa74,Arn82,JoPi86,BoMaHe91,FaRi99,HuHuRa99,CoKaSh00,GiHeWa02,RiWh03,MoRi05,GrScSc06,AiMoMu06,BeLaPi06,DuKa06,DeSeWh08,PuAmKa09,WiStBuGh10,DeSe10,EtChViGl10},
with optimized efficiency because of its tensorized basis functions
\citep{WiStBuGh10,AcKo11}. In particular, it can accurately handle
very distorted mesh elements \citep{OlSe11}.\\

It has very good accuracy and convergence properties \citep{MaPa89,SePr94,DeFiMu02,Coh02,DeSe07,SeOl08,AiWa09,AiWa10,MeStTh12}.
The spectral element approach admits spectral rates of convergence
and allows exploiting $hp$-convergence schemes. It is also very well
suited to parallel implementation on very large supercomputers \citep{KoTsChTr03,TsKoChTr03,KoLaMi08a,CaKoLaTiMiLeSnTr08,KoViCh10}
as well as on clusters of GPU accelerating graphics cards \citep{Kom11,MiKo10,KoMiEr09,KoErGoMi10}.
Tensor products inside each element can be optimized to reach very
high efficiency \citep{DeFiMu02}, and mesh point and element numbering
can be optimized to reduce processor cache misses and improve cache
reuse \citep{KoLaMi08a}. The SEM can also handle triangular (in 2D)
or tetrahedral (in 3D) elements \citep{WinBoyd96,TaWi00,KoMaTrTaWi01,Coh02,MeViSa06}
as well as mixed meshes, although with increased cost and reduced
accuracy in these elements, as in the discontinuous Galerkin method.\\

Note that in many geological models in the context of seismic wave
propagation studies (except for instance for fault dynamic rupture
studies, in which very high frequencies or supershear rupture need
to be modeled near the fault, see e.g. \citet{BeGlCrViPi07,BeGlCrVi09,PuAmKa09,TaCrEtViBeSa10})
a continuous formulation is sufficient because material property contrasts
are not drastic and thus conforming mesh doubling bricks can efficiently
handle mesh size variations \citep{KoTr02a,KoLiTrSuStSh04,LeChLiKoHuTr08,LeChKoHuTr09,LeKoHuTr09}.\\

For a detailed introduction to the SEM as applied to regional seismic
wave propagation, please consult \citet{PeKoLuMaLeCaLeMaLiBlNiBaTr11,TrKoLi08,KoVi98,KoTr99,ChKoViCaVaFe07}
and in particular \citet{LeKoHuTr09,LeChKoHuTr09,LeChLiKoHuTr08,GoAmTaCaSmSaMaKo09,WiKoScTr04,KoLiTrSuStSh04}.
A detailed theoretical analysis of the dispersion
and stability properties of the SEM is available in \citet{Coh02}, \citet{DeSe07}, \citet{SeOl07}, \citet{SeOl08} and \citet{MeStTh12}.\\

Effects due to lateral variations in compressional-wave speed, shear-wave
speed, density, a 3D crustal model, topography and bathymetry are
included. The package can accommodate full 21-parameter anisotropy
(see~\citet{ChTr07}) as well as lateral variations in attenuation
\citep{SaKoTr10}. Adjoint capabilities and finite-frequency kernel
simulations are included \citep{TrKoLi08,PeKoLuMaLeCaLeMaLiBlNiBaTr11,LiTr06,FiIgBuKe09,ViOp09}.\\

The SEM was originally developed in computational fluid dynamics \citep{Pat84,MaPa89}
and has been successfully adapted to address problems in seismic wave
propagation. Early seismic wave propagation applications of the SEM,
utilizing Legendre basis functions and a perfectly diagonal mass matrix,
include \citet{CoJoTo93}, \citet{Kom97}, \citet{FaMaPaQu97}, \citet{CaGa97},
\citet{KoVi98} and \citet{KoTr99}, whereas applications involving
Chebyshev basis functions and a nondiagonal mass matrix include \citet{SePr94},
\citet{PrCaSe94} and \citet{SePrPr95}.\\

All SPECFEM3D software is written in Fortran2003 with full portability
in mind, and conforms strictly to the Fortran2003 standard. It uses
no obsolete or obsolescent features of Fortran. The package uses parallel
programming based upon the Message Passing Interface (MPI) \citep{GrLuSk94,Pac97}.\\

SPECFEM3D won the Gordon Bell award for best performance at the SuperComputing~2003
conference in Phoenix, Arizona (USA) (see \citet{KoTsChTr03} and
\url{www.sc-conference.org/sc2003/nrfinalaward.html}). It was a finalist
again in 2008 for a run at 0.16 petaflops (sustained) on 149,784 processors
of the `Jaguar' Cray XT5 system at Oak Ridge National Laboratories
(USA) \citep{CaKoLaTiMiLeSnTr08}. It also won the BULL Joseph Fourier
supercomputing award in 2010.\\

It reached the sustained one petaflop performance level for the first time in February 2013
on the Blue Waters Cray supercomputer at the National Center for Supercomputing Applications (NCSA), located at the University of Illinois at Urbana-Champaign (USA).\\

This new release of the code includes Convolution or
Auxiliary Differential Equation Perfectly Matched absorbing Layers
(C-PML or ADE-PML) \citep{MaKoEz08,MaKoGe08,MaKo09,MaKoGeBr10,KoMa07}.
It also includes support for GPU graphics card acceleration \citep{Kom11,MiKo10,KoMiEr09,KoErGoMi10}.\\

The next release of the code will use the PT-SCOTCH parallel and threaded
version of SCOTCH for mesh partitioning instead of the serial version.\\

SPECFEM3D Cartesian includes coupled fluid-solid domains and adjoint
capabilities, which enables one to address seismological inverse problems,
but for linear rheologies only so far. To accommodate visco-plastic
or non-linear rheologies, the reader can refer to the \href{http://geoelse.stru.polimi.it/}{GeoELSE}software
package \citep{CaGa97,StPaIg09}.

\section{Citation}

If you use SPECFEM3D Cartesian for your own research, please cite
at least one of the following articles:
\begin{description}
\item [{Numerical simulations in general}] ~\\
 Forward and adjoint simulations are described in detail in \citet{TrKoLi08,PeKoLuMaLeCaLeMaLiBlNiBaTr11,VaCaSaKoVi99,KoMiEr09,KoErGoMi10,ChKoViCaVaFe07,MaKoDi09,KoViCh10,CaKoLaTiMiLeSnTr08,TrKoHjLiZhPeBoMcFrTrHu10,KoRiTr02,KoTr02a,KoTr02b,KoTr99}
or \citet{KoVi98}. Additional aspects of adjoint simulations are
described in \citet{TrTaLi05,LiTr06,TrKoLi08,LiTr08,TrKoHjLiZhPeBoMcFrTrHu10,PeKoLuMaLeCaLeMaLiBlNiBaTr11}.
Domain decomposition is explained in detail in \citet{MaKoBlLe08},
and excellent scaling up to 150,000 processor cores is shown for instance
in \citet{CaKoLaTiMiLeSnTr08,KoLaMi08a,MaKoBlLe08,KoErGoMi10,Kom11},
\item [{GPU computing}] ~\\
 Computing on GPU graphics cards for acoustic or seismic wave propagation
applications is described in detail in \citet{Kom11,MiKo10,KoMiEr09,KoErGoMi10}.
\end{description}
\noindent If you use this new version, which has non blocking
MPI for much better performance for medium or large runs, please cite
at least one of these six articles, in which results of non blocking
MPI runs are presented: \citet{PeKoLuMaLeCaLeMaLiBlNiBaTr11,KoErGoMi10,KoViCh10,Kom11,CaKoLaTiMiLeSnTr08,MaKoBlLe08}.\\

\noindent If you work on geophysical applications, you may be interested
in citing some of these application articles as well, among others:
\begin{description}
\item [{Southern California simulations}] ~\\
 \citet{KoLiTrSuStSh04,KrJiKoTr06a,KrJiKoTr06b}.

If you use the 3D southern California model, please cite \citet{SuSh03}
(Los Angeles model), \citet{lovelyetal06} (Salton Trough), and \citet{hauksson2000}
(southern California). The Moho map was determined by \citet{zhu&kanamori2000}.
The 1D SoCal model was developed by \citet{DrHe90}.

\item [{Anisotropy}] ~\\
 \citet{ChTr07,JiTsKoTr05,ChFaKo04,FaChKo04,RiRiKoTrHe02,TrKo00}.
\item [{Attenuation}] ~\\
 \citet{SaKoTr10,KoTr02a,KoTr99}.
\item [{Topography}] ~\\
 \citet{LeKoHuTr09,LeChKoHuTr09,LeChLiKoHuTr08,GoAmTaCaSmSaMaKo09,WiKoScTr04}.
\end{description}
The corresponding Bib\TeX{} entries may be found in file \texttt{doc/USER\_MANUAL/bibliography.bib}.

\section{Support}

This material is based upon work supported by the USA National Science
Foundation under Grants No. EAR-0406751 and EAR-0711177, by the French
CNRS, French INRIA Sud-Ouest MAGIQUE-3D, French ANR NUMASIS under
Grant No. ANR-05-CIGC-002, and European FP6 Marie Curie International
Reintegration Grant No. MIRG-CT-2005-017461. Any opinions, findings,
and conclusions or recommendations expressed in this material are
those of the authors and do not necessarily reflect the views of the
USA National Science Foundation, CNRS, INRIA, ANR or the European
Marie Curie program.

\chapter{\label{cha:Getting-Started}Getting Started}

To download the SPECFEM3D\_Cartesian software package, type this:
\begin{lyxcode}
git clone --recursive --branch devel https://github.com/geodynamics/specfem3d.git
\end{lyxcode}
{\small \par}

We recommend that you add {\texttt{ulimit -S -s unlimited}} to your
{\texttt{.bash\_profile}} file and/or {\texttt{limit stacksize
unlimited }} to your {\texttt{.cshrc}} file to suppress any potential
limit to the size of the Unix stack.\\

Then, to configure the software for your system, run the \texttt{configure}
shell script. This script will attempt to guess the appropriate configuration
values for your system. However, at a minimum, it is recommended that
you explicitly specify the appropriate command names for your Fortran
compiler and MPI package (another option is to define FC, CC and MPIF90
in your .bash\_profile or your .cshrc file):
\begin{lyxcode}
./configure~FC=ifort~MPIFC=mpif90
\end{lyxcode}
Note that MPI must be installed with MPI-IO enabled because parts of SPECFEM3D perform I/Os through MPI-IO.

Before running the \texttt{configure} script, you should probably
edit file \texttt{flags.guess} to make sure that it contains the best
compiler options for your system. Known issues or things to check
are:
\begin{description}
\item [{\texttt{Intel ifort compiler}}] See if you need to add \texttt{-assume byterecl} for your machine. \textbf{In the case of that compiler, we have noticed that versions dot zero sometimes have bugs or issues that can lead to wrong results when running the code, thus we \emph{strongly} recommend using versions dot one or above (for instance version 13.1 instead of 13.0, version 14.1 instead of 14.0 and so on)}.
\item [{\texttt{IBM compiler}}] See if you need to add \texttt{-qsave}
or \texttt{-qnosave} for your machine.
\item [{\texttt{Mac OS}}] You will probably need to install \texttt{XCODE}.
\end{description}
When compiling on an IBM machine with the \texttt{xlf} and \texttt{xlc}
compilers, we suggest running the \texttt{configure} script with the
following options:\\

\noindent \texttt{./configure FC=xlf90\_r MPIFC=mpif90 CC=xlc\_r CFLAGS=\textquotedbl{}-O3
-q64\textquotedbl{} FCFLAGS=\textquotedbl{}-O3 -q64\textquotedbl{}
-with-scotch-dir=...}.\\

On SGI systems, \texttt{flags.guess} automatically informs \texttt{configure}
to insert `\texttt{`TRAP\_FPE=OFF}'' into the generated \texttt{Makefile}
in order to turn underflow trapping off.\\

You can add \texttt{--enable-vectorization} to the configuration options to speed up the code in the fluid (acoustic) and elastic parts.
This works fine if (and only if) your computer always allocates a contiguous memory block for each allocatable array;
this is the case for most machines and most compilers, but not all. To disable this feature, use option \texttt{--disable-vectorization}.
For more details see \href{https://github.com/geodynamics/specfem3d/issues/81}{github.com/geodynamics/specfem3d/issues/81} .
To check if that option works fine on your machine, run the code with and without it for an acoustic/elastic model and make sure the seismograms are identical.\\

Note that we use CUBIT (now called Trelis) to create meshes of hexahedra, but other packages
can be used as well, for instance GiD from \url{http://gid.cimne.upc.es}
or Gmsh from \url{http://geuz.org/gmsh} \citep{GeRe09}. Even mesh
creation packages that generate tetrahedra, for instance TetGen from
\url{http://tetgen.berlios.de}, can be used because each tetrahedron
can then easily be decomposed into four hexahedra as shown in the
picture of the TetGen logo at \url{http://tetgen.berlios.de/figs/Delaunay-Voronoi-3D.gif};
while this approach does not generate hexahedra of optimal quality,
it can ease mesh creation in some situations and it has been shown
that the spectral-element method can very accurately handle distorted
mesh elements \citep{OlSe11}.

The SPECFEM3D Cartesian software package relies on the SCOTCH library
to partition meshes created with CUBIT. METIS \citep{KaKu98a,KaKu98c,KaKu98b}
can also be used instead of SCOTCH if you prefer, by editing file
src/decompose\_mesh/decompose\_mesh.F90 and uncommenting the flag
USE\_METIS\_INSTEAD\_OF\_SCOTCH. You will also then need to install
and compile Metis version 4.0 (do {*}NOT{*} install Metis version
5.0, which has incompatible function calls) and edit \texttt{src/decompose\_mesh/Makefile.in}
and uncomment the METIS link flag in that file before running \texttt{configure}.\\

The SCOTCH library \citep{PeRo96} provides efficent static mapping,
graph and mesh partitioning routines. SCOTCH is a free software package
developed by Fran\c{c}ois Pellegrini et al. from LaBRI and INRIA in Bordeaux,
France, downloadable from the web page \url{https://gforge.inria.fr/projects/scotch/}.
In case no SCOTCH libraries can be found on the system, the configuration
will bundle the version provided with the source code for compilation.
The path to an existing SCOTCH installation can to be set explicitly
with the option \texttt{-{}-with-scotch-dir}. Just as an example:
\begin{lyxcode}
./configure~FC=ifort~MPIFC=mpif90~-{}-with-scotch-dir=/opt/scotch
\end{lyxcode}
If you use the Intel ifort compiler to compile the code, we recommend
that you use the Intel icc C compiler to compile Scotch, i.e., use:
\begin{lyxcode}
./configure~CC=icc~FC=ifort~MPIFC=mpif90
\end{lyxcode}

When compiling the SCOTCH source code, if you get a message such as: "ld: cannot find -lz",
the Zlib compression development library is probably missing on your machine and you will need to install it or ask your system administrator to
do so. On Linux machines the package is often called "zlib1g-dev" or similar. (thus "sudo apt-get install zlib1g-dev" would install it)

To compile a serial version of the code for small meshes that fits
on one compute node and can therefore be run serially, run \texttt{configure}
with the \texttt{-{}-without-mpi} option to suppress all calls to
MPI.

A summary of the most important configuration variables follows.
\begin{description}
\item [{\texttt{F90}}] Path to the Fortran compiler.
\item [{\texttt{MPIF90}}] Path to MPI Fortran.
\item [{\texttt{MPI\_FLAGS}}] Some systems require this flag to link to
MPI libraries.
\item [{\texttt{FLAGS\_CHECK}}] Compiler flags.
\end{description}
The configuration script automatically creates for each executable
a corresponding \texttt{Makefile} in the \texttt{src/} subdirectoy.
The \texttt{Makefile} contains a number of suggested entries for various
compilers, e.g., Portland, Intel, Absoft, NAG, and Lahey. The software
has run on a wide variety of compute platforms, e.g., various PC clusters
and machines from Sun, SGI, IBM, Compaq, and NEC. Select the compiler
you wish to use on your system and choose the related optimization
flags. Note that the default flags in the \texttt{Makefile} are undoubtedly
not optimal for your system, so we encourage you to experiment with
these flags and to solicit advice from your systems administrator.
Selecting the right compiler and optimization flags can make a tremendous
difference in terms of performance. We welcome feedback on your experience
with various compilers and flags.

Now that you have set the compiler information, you need to select
a number of flags in the \texttt{constants.h} file depending on your
system:
\begin{description}
\item [{\texttt{LOCAL\_PATH\_IS\_ALSO\_GLOBAL}}] Set to \texttt{.false.}
on most cluster applications. For reasons of speed, the (parallel)
distributed database generator typically writes a (parallel) database
for the solver on the local disks of the compute nodes. Some systems
have no local disks, e.g., BlueGene or the Earth Simulator, and other
systems have a fast parallel file system, in which case this flag
should be set to \texttt{.true.}. Note that this flag is not used
by the database generator or the solver; it is only used for some
of the post-processing.
\end{description}
The package can run either in single or in double precision mode.
The default is single precision because for almost all calculations
performed using the spectral-element method using single precision
is sufficient and gives the same results (i.e. the same seismograms);
and the single precision code is faster and requires exactly half
as much memory. Select your preference by selecting the appropriate
setting in the \texttt{constants.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_REAL}}] Set to \texttt{SIZE\_REAL} for single precision
and \texttt{SIZE\_DOUBLE} for double precision.
\end{description}
In the \texttt{precision.h} file:
\begin{description}
\item [{\texttt{CUSTOM\_MPI\_TYPE}}] Set to \texttt{MPI\_REAL} for single
precision and \texttt{MPI\_DOUBLE\_PRECISION} for double precision.
\end{description}
On many current processors (e.g., Intel, AMD, IBM Power), single precision
calculations are significantly faster; the difference can typically
be 10\% to 25\%. It is therefore better to use single precision. What
you can do once for the physical problem you want to study is run
the same calculation in single precision and in double precision on
your system and compare the seismograms. If they are identical (and
in most cases they will), you can select single precision for your
future runs.\\

If your compiler has problems with the \texttt{use mpi} statements that are used in the code, use the script called
\texttt{replace\_use\_mpi\_with\_include\_mpif\_dot\_h.pl} in the root directory to replace all of them with \texttt{include 'mpif.h'} automatically.

\section{Adding OpenMP support in addition to MPI}

NOTE FROM JULY 2013: OpenMP support is maybe / probably not maintained any more.
Thus the section below is maybe obsolete.

OpenMP support can be enabled in addition to MPI. However, in many
cases performance will not improve because our pure MPI implementation
is already heavily optimized and thus the resulting code will in fact
be slightly slower. A possible exception could be IBM BlueGene-type
architectures.

To enable OpenMP, uncomment the OpenMP compiler option in two lines
in file \texttt{src/specfem3D/Makefile.in} (before running \texttt{configure})
and also uncomment the \texttt{\#define USE\_OPENMP} statement in
file\\
 \texttt{src/specfem3D/specfem3D.F90}.

The DO-loop using OpenMP threads has a SCHEDULE property. The OMP\_SCHEDULE
environment variable can set the scheduling policy of that DO-loop.
Tests performed by Marcin Zielinski at SARA (The Netherlands) showed
that often the best scheduling policy is DYNAMIC with the size of
the chunk equal to the number of OpenMP threads, but most preferably
being twice as the number of OpenMP threads (thus chunk size = 8 for
4 OpenMP threads etc). If OMP\_SCHEDULE is not set or is empty, the
DO-loop will assume generic scheduling policy, which will slow down
the job quite a bit.

\section{Compiling on an IBM BlueGene}

\underline{More recent installation instruction for IBM BlueGene, from April 2013:}\\

\noindent Edit file \texttt{flags.guess} and put this for \texttt{FLAGS\_CHECK}:\\

\noindent \texttt{-g -qfullpath -O2 -qsave -qstrict -qtune=qp -qarch=qp
-qcache=auto -qhalt=w}\\
 \texttt{-qfree=f90 -qsuffix=f=f90 -qlanglvl=95pure -Q -Q+rank,swap\_all
-Wl,-relax}\\

\noindent The most relevant are the -qarch and -qtune flags, otherwise
if these flags are set to ``auto'' then they are wrongly assigned
to the architecture of the frond-end node, which is different from
that on the compute nodes. You will need to set these flags to the
right architecture for your BlueGene compute nodes, which is not necessarily
``qp''; ask your system administrator. On some machines if is necessary
to use -O2 in these flags instead of -O3 due to a compiler bug of
the XLF version installed. We thus suggest to first try -O3, and then
if the code does not compile or does not run fine then switch back
to -O2. The debug flags (-g, -qfullpath) do not influence performance
but are useful to get at least some insights in case of problems.\\

\noindent Before running \texttt{configure}, select the XL Fortran
compiler by typing \texttt{module load bgq-xl/1.0} or \texttt{module
load bgq-xl} (another, less efficient option is to load the GNU compilers
using \texttt{module load bgq-gnu/4.4.6} or similar).\\

\noindent Then, to configure the code, type this:\\

\noindent \texttt{./configure FC=bgxlf90\_r MPIFC=mpixlf90\_r CC=bgxlc\_r
LOCAL\_PATH\_IS\_ALSO\_GLOBAL=true}\\

In order for the SCOTCH domain decomposer to compile, on some (but
not all) Blue Gene systems you may need to run \texttt{configure}
with \texttt{CC=gcc} instead of \texttt{CC=bgxlc\_r}.

\noindent \underline{Older installation instruction for IBM BlueGene, from 2011:}\\

\noindent To compile the code on an IBM BlueGene, Laurent L\'eger from
IDRIS, France, suggests the following: compile the code with

\noindent \texttt{ FLAGS\_CHECK=\textquotedbl{}-O3 -qsave -qstrict
-qtune=auto -qarch=450d -qcache=auto }~\\
 \texttt{ -qfree=f90 -qsuffix=f=f90 -g -qlanglvl=95pure -qhalt=w -Q
-Q+rank,swap\_all -Wl,-relax\textquotedbl{}}

\noindent Option \textquotedbl{}-Wl,-relax\textquotedbl{} must be
added on many (but not all) BlueGene systems to be able to link the
binaries \texttt{xmeshfem3D} and \texttt{xspecfem3D} because the final
link step is done by the GNU \texttt{ld} linker even if one uses \texttt{FC=bgxlf90\_r,
MPIFC=mpixlf90\_r} and \texttt{CC=bgxlc\_r} to create all the object
files. On the contrary, on some BlueGene systems that use the native
AIX linker option \textquotedbl{}-Wl,-relax\textquotedbl{} can lead
to problems and must be suppressed from \texttt{flags.guess}.

\noindent Also, \texttt{AR=ar, ARFLAGS=cru} and \texttt{RANLIB=ranlib}
are hardwired in all \texttt{Makefile.in} files by default, but to
cross-compile on BlueGene/P one needs to change these values to \texttt{AR=bgar,
ARFLAGS=cru} and \texttt{RANLIB=bgranlib}. Thus the easiest thing
to do is to modify all \texttt{Makefile.in} files and the \texttt{configure}
script to set them automatically by \texttt{configure}. One then just
needs to pass the right commands to the \texttt{configure} script:

>>>>

\chapter{Running through a Scheduler}\label{cha:Scheduler}
